\textit{Supplementary materials for}: Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-01-23 paper/boneyard.tex

\subsection*{Context and the structure of memory}
Remembering words on a studied list requires distinguishing the current list from the rest of one's experience.  To model this fundamental memory capability, cognitive scientists have posited the existence of a special representation, called \emph{context}, that is associated with each list.  According to early theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are composed of many features which fluctuate from moment to moment, slowly drifting through a multidimensional feature space.  During recall, this representation forms part of the retrieval cue, enabling us to distinguish list items from non-list items.  Understanding the role of context in memory processes is particularly important in self-cued memory tasks, such as \textit{free recall}, where the retrieval cue is ``context'' itself.  In free recall, participants study lists of items and are instructed to recall the items in any order they choose.

@@

Our experimental paradigm incorporates the Google Cloud Speech API speech-to-text engine to automatically transcribe participants' verbal recalls into text.  This allows recalls to be transcribed in real time-- a distinguishing feature of the experiment that is critical for using participants' recalls of prior lists to adapt the presentation orders of future lists, as in Experiments 2 and 3.  (In typical verbal recall experiments the audio data must be parsed manually, precluding this sort of design.)  We recently published a study verifying that the computer-generated recall transcripts are sufficiently similar to human-generated transcripts to yield reliable data~\citep{ZimaEtal18}.

\subsection*{Experiment 1: tracking the dynamics of memory organization over successive learning trials}
% maybe this could benefit from reframing wrt to the variance idea?
Large portions of the memory theory and education literatures have focused on identifying learners' unique learning styles.  In the memory theory literature~\citep{SmitVela01, Kaha12, HealKaha14a, MannEtal15}, the focus has been on identifying behavioral signatures of how memories are organized (i.e., in the language of this proposal, identifying memory fingerprints).  In the education literature, the focus has been been on asking whether catering lessons to an individual's unique learning style can help them to learn more efficiently~\citep{PashEtal08}.  Each of these literatures effectively treats people's learning styles as a fixed point.  \textbf{In other words, the assumption underlying the majority of these studies is that each individual has their own unique learning style, and that learning style remains roughly consistent over time.  Experiment 1 is designed to explicitly test this assumption using the feature-rich free recall paradigm (Fig.~\ref{fig:exp1}).}  Importantly, we will collect data from two populations: (a) members of the Dartmouth and surrounding community, with the experiment running in a tightly controlled laboratory setting, and (b) Mechanical Turk workers, via the Internet.  Taken together, we hope that these two populations of participants will allow us to generalize our findings to a diverse set of learners.

\paragraph{Design and procedure.}
We will use the feature-rich free recall paradigm (Fig.~\ref{fig:exp1}) to study how people's memory fingerprints are influenced by presentation order.  The experiment will be divided into six manipulation conditions (one for each stimulus feature); in each of these manipulation conditions the words from the first 8 lists participants study will be sorted according to one of the stimulus dimensions.  For example, in the ``color'' condition, words from the first 8 lists will be sorted by font color; in the ``starting letter'' condition, words from the first 8 lists will be sorted alphabetically; etc.  The second 8 lists will be presented in a random order in every condition.  As described below, we will assess how participants' memories are affected by these order manipulations, both for their memories of words on first 8 lists where the manipulations are applied, as well as on the second 8 lists.  Participants' behaviors will be compared to a seventh (control) condition, where the presentation order is randomized for all 16 lists.

@@

\paragraph{Key analyses.}
Experiment 1 is motivated by two fundamental questions about the dynamics of memory organization and recall.  \textbf{First, we want to understand how memory fingerprints vary across people and over time.}  For example, are people's memory fingerprints generally stable across lists?  Or, what are the similarities and differences across people (in how people organize their memories, and how their memory fingerprints change over time)?  Are those individual differences predictive of memory performance? \textbf{Second, we want to understand how people's memory fingerprints can be manipulated.}  Specifically, if lists are sorted according to a particular feature (e.g. the words' semantic categories), how does that influence people's tendencies to cluster their recalls of those words with respect to that feature?  And which of those tendencies persist during recalls of later (randomly ordered) lists?  For example, if we organize early lists alphabetically, are  participants more likely recall the words from those lists alphabetically as well?  More generally, might there be some way of influencing how people organize their memories, even after the influencing cue is no longer present?

@@

\paragraph{Experiment 1 preliminary data.}
We ran a preliminary study ($n = 172$) to measure participants' memory fingerprints under the in-laboratory conditions outlined in Experiment 1.  Using data from the control condition (where all lists were ordered randomly; $n = 45$), we computed the distribution of average memory fingerprints (across lists) from each participant (Fig.~\ref{fig:random_pilot}a).  The degree to which participants organize their memories with respect to each feature varies substantially across people (e.g.\ there is a wide spread of values covered across participants within each feature dimension).  In addition, some features are more likely to be leveraged in people's memories than others.  For example, participants are more likely to organize their recalls by category, size, and temporal proximity than by font color or word length.  We also examined how stable participants' memory fingerprints were across lists.  We found that participants whose memory fingerprints were more stable across lists (i.e., lower variance) remembered more words (Fig.~\ref{fig:random_pilot}b).  We also examined the manipulation conditions, where the first 8 lists participants studied were sorted according to one feature (category: $n=21$, color: $n=20$, screen location: $n=21$, first letter: $n=20$, word length: $n=23$, size: $n=22$).  Relative to randomly sorted lists, sorting by some features increased the average memory fingerprint variability and sorting by other features decreased the average memory fingerprint variability.  (We used a permutation-based procedure to compute the variability, which controlled for the distributions of feature values of the words participants remembered.)  We found that participants remembered more words in the conditions where the presentation orders led to lower fingerprint variability (Fig.~\ref{fig:random_pilot}c).  Taken together, these preliminary data suggest that participants who naturally exhibit more stable fingerprints remember more words, and fingerprint stability may be influenced by the order in which words are presented.  In turn, sorting the word presentations along different dimensions can enhance (or diminish) memory performance according to whether fingerprint variability is increased or decreased, respectively.

@@

\subsection*{Experiment 3: adaptive list learning using (neural) memory fingerprints}
The above experiments leverage participants' \textit{behaviors} (i.e., the sequences of recalls they make) to optimize learning by changing the order in which subsequent lists are presented.  In Experiment 3, we seek to optimize learning using \textit{neural} (EEG) data. \textbf{Our prior work shows that the neural patterns recorded while participants study a list of words may be used to predict their individual tendencies to cluster their recalls according to presentation order (temporal clustering; Fig.~\ref{fig:clust}a) or word meaning (semantic clustering; Fig.~\ref{fig:clust}b).}

@@

\paragraph{Design and procedure.} The experiment and parameters will be analogous to Experiment 2, except that we will use EEG data recorded as participants are studying the words (rather than behavioral data from participants' recalls) to estimate their memory fingerprints.  We plan to record the EEG data using an ActiChamp 160-electrode system (with a photosensor-based electrode mapping system, to allow us to perform source-localization analyses).  Participants will study a total of 16 lists of 16 words each; we will adaptively adjust the presentation orders of each list as described next.

>>>>

\subsection*{Context and the structure of memory}

Remembering words on a studied list requires distinguishing the current list
from the rest of one's experience. To model this fundamental memory capability,
cognitive scientists have posited the existence of a special representation,
called \emph{context}, that is associated with each list. According to early
theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are
composed of many features which fluctuate from moment to moment, slowly
drifting through a multidimensional feature space. During recall, this
representation forms part of the retrieval cue, enabling us to distinguish list
items from non-list items. Understanding the role of context in memory
processes is particularly important in self-cued memory tasks, such as
\textit{free recall}, where the retrieval cue is ``context'' itself. In free
recall, participants study lists of items and are instructed to recall the
items in any order they choose.

@@

Our experimental paradigm incorporates the Google Cloud Speech API
speech-to-text engine to automatically transcribe participants' verbal recalls
into text. This allows recalls to be transcribed in real time-- a
distinguishing feature of the experiment that is critical for using
participants' recalls of prior lists to adapt the presentation orders of future
lists, as in Experiments 2 and 3. (In typical verbal recall experiments the
audio data must be parsed manually, precluding this sort of design.) We
recently published a study verifying that the computer-generated recall
transcripts are sufficiently similar to human-generated transcripts to yield
reliable data~\citep{ZimaEtal18}. \subsection*{Experiment 1: tracking the
dynamics of memory organization over successive learning trials} % maybe this
could benefit from reframing wrt to the variance idea? Large portions of the
memory theory and education literatures have focused on identifying learners'
unique learning styles. In the memory theory literature~\citep{SmitVela01,
Kaha12, HealKaha14a, MannEtal15}, the focus has been on identifying behavioral
signatures of how memories are organized (i.e., in the language of this
proposal, identifying memory fingerprints). In the education literature, the
focus has been been on asking whether catering lessons to an individual's
unique learning style can help them to learn more
efficiently~\citep{PashEtal08}. Each of these literatures effectively treats
people's learning styles as a fixed point. \textbf{In other words, the
assumption underlying the majority of these studies is that each individual has
their own unique learning style, and that learning style remains roughly
consistent over time. Experiment 1 is designed to explicitly test this
assumption using the feature-rich free recall paradigm (Fig.~\ref{fig:exp1}).}
Importantly, we will collect data from two populations: (a) members of the
Dartmouth and surrounding community, with the experiment running in a tightly
controlled laboratory setting, and (b) Mechanical Turk workers, via the
Internet. Taken together, we hope that these two populations of participants
will allow us to generalize our findings to a diverse set of learners.

\paragraph{Design and procedure.}

We will use the feature-rich free recall paradigm (Fig.~\ref{fig:exp1}) to
study how people's memory fingerprints are influenced by presentation order.
The experiment will be divided into six manipulation conditions (one for each
stimulus feature); in each of these manipulation conditions the words from the
first 8 lists participants study will be sorted according to one of the
stimulus dimensions. For example, in the ``color'' condition, words from the
first 8 lists will be sorted by font color; in the ``starting letter''
condition, words from the first 8 lists will be sorted alphabetically; etc. The
second 8 lists will be presented in a random order in every condition. As
described below, we will assess how participants' memories are affected by
these order manipulations, both for their memories of words on first 8 lists
where the manipulations are applied, as well as on the second 8 lists.
Participants' behaviors will be compared to a seventh (control) condition,
where the presentation order is randomized for all 16 lists.

@@

\paragraph{Key analyses.}

Experiment 1 is motivated by two fundamental questions about the dynamics of
memory organization and recall. \textbf{First, we want to understand how memory
fingerprints vary across people and over time.} For example, are people's
memory fingerprints generally stable across lists? Or, what are the
similarities and differences across people (in how people organize their
memories, and how their memory fingerprints change over time)? Are those
individual differences predictive of memory performance? \textbf{Second, we
want to understand how people's memory fingerprints can be manipulated.}
Specifically, if lists are sorted according to a particular feature (e.g. the
words' semantic categories), how does that influence people's tendencies to
cluster their recalls of those words with respect to that feature? And which of
those tendencies persist during recalls of later (randomly ordered) lists? For
example, if we organize early lists alphabetically, are participants more
likely recall the words from those lists alphabetically as well? More
generally, might there be some way of influencing how people organize their
memories, even after the influencing cue is no longer present?

@@

\paragraph{Experiment 1 preliminary data.}

We ran a preliminary study ($n = 172$) to measure participants' memory
fingerprints under the in-laboratory conditions outlined in Experiment 1. Using
data from the control condition (where all lists were ordered randomly; $n =
45$), we computed the distribution of average memory fingerprints (across
lists) from each participant (Fig.~\ref{fig:random_pilot}a). The degree to
which participants organize their memories with respect to each feature varies
substantially across people (e.g.\ there is a wide spread of values covered
across participants within each feature dimension). In addition, some features
are more likely to be leveraged in people's memories than others. For example,
participants are more likely to organize their recalls by category, size, and
temporal proximity than by font color or word length. We also examined how
stable participants' memory fingerprints were across lists. We found that
participants whose memory fingerprints were more stable across lists (i.e.,
lower variance) remembered more words (Fig.~\ref{fig:random_pilot}b). We also
examined the manipulation conditions, where the first 8 lists participants
studied were sorted according to one feature (category: $n=21$, color: $n=20$,
screen location: $n=21$, first letter: $n=20$, word length: $n=23$, size:
$n=22$). Relative to randomly sorted lists, sorting by some features increased
the average memory fingerprint variability and sorting by other features
decreased the average memory fingerprint variability. (We used a
permutation-based procedure to compute the variability, which controlled for
the distributions of feature values of the words participants remembered.) We
found that participants remembered more words in the conditions where the
presentation orders led to lower fingerprint variability
(Fig.~\ref{fig:random_pilot}c). Taken together, these preliminary data suggest
that participants who naturally exhibit more stable fingerprints remember more
words, and fingerprint stability may be influenced by the order in which words
are presented. In turn, sorting the word presentations along different
dimensions can enhance (or diminish) memory performance according to whether
fingerprint variability is increased or decreased, respectively.

@@

\subsection*{Experiment 3: adaptive list learning using (neural) memory fingerprints}

The above experiments leverage participants' \textit{behaviors} (i.e., the
sequences of recalls they make) to optimize learning by changing the order in
which subsequent lists are presented. In Experiment 3, we seek to optimize
learning using \textit{neural} (EEG) data. \textbf{Our prior work shows that
the neural patterns recorded while participants study a list of words may be
used to predict their individual tendencies to cluster their recalls according
to presentation order (temporal clustering; Fig.~\ref{fig:clust}a) or word
meaning (semantic clustering; Fig.~\ref{fig:clust}b).}

@@

\paragraph{Design and procedure.}

The experiment and parameters will be analogous to Experiment 2, except that we
will use EEG data recorded as participants are studying the words (rather than
behavioral data from participants' recalls) to estimate their memory
fingerprints. We plan to record the EEG data using an ActiChamp 160-electrode
system (with a photosensor-based electrode mapping system, to allow us to
perform source-localization analyses). Participants will study a total of 16
lists of 16 words each; we will adaptively adjust the presentation orders of
each list as described next.
